Liveness-Based Garbage Collection for Lazy Languages

### 2016-04-28 paper/paper.tex
replaced While with Whereas as suggested

\begin{abstract} 
We consider  the problem of reducing  the memory required to  run lazy
first-order functional programs. Our approach is to analyze
programs  for liveness  of  heap-allocated data.   The  result of  the
analysis is  used to preserve  only live data---a subset  of reachable
data---during garbage  collection.  The result  is an increase  in the
garbage reclaimed and  a  reduction in  the peak memory requirement 
of programs.   While this  technique has already  been shown  to yield
benefits for  eager first-order  languages, the  lack of  a statically
determinable execution  order and  the presence  of closures  pose new
challenges  for lazy  languages.  These  require changes  both in  the
liveness analysis itself and in the design of the garbage collector.

@@

While  the runtime  system  of most  functional  languages includes  a
garbage collector to efficiently  reclaim memory, empirical studies on
Scheme~\cite{karkare06effectiveness}  and  Haskell~\cite{rojemo96lag}
programs have shown that garbage  collectors leave uncollected a large
number of  memory objects that are  reachable but not live  (here {\em
  live} means the  object may potentially be used by  the program at a
later stage).  This results in unnecessary memory retention.

@@

While  the idea  of  using static  analysis  to improve  memory
  utilization  has  been  shown  to   be  effective  for  {\em  eager}
  languages~\cite{asati14lgc,       HofmannJ03,       inoue88analysis,
    lee05static}, a straightforward extension  of the technique is not
  possible  for  lazy  languages,  where  heap-allocated  objects  may
  include   closures    (runtime   representations    of   unevaluated
  expressions).
%%\todelete{Designing  a  liveness  based  garbage
%%  collector   for  lazy   languages  poses   significant  challenges.}
Firstly, since  data is made  live by  evaluation of closures,  and in
lazy languages  the place in  the program where this  evaluation takes
place cannot  be statically determined, laziness  complicates liveness
analysis itself.  Moreover, for liveness-based  GC to be effective, we
need to  extend it  to closures  apart from  evaluated data.   Since a
closure can escape the scope in which  it was created, it has to carry
the  liveness  information  of   its  free  variables.   As  execution
progresses  and possible  future uses  are eliminated,  we update  the
liveness  information  in  a  closure with  a  more  precise  version.
For these  reasons,  the  garbage   collector  also  becomes
significantly more  complicated than a liveness-based  collector for an
eager language.

@@

Section~\ref{sec:defs}  introduces  the   syntax  of  the  programming
language considered  and gives a small-step  operational semantics for
it.  The liveness  analysis for this language and  its soundness proof
is          presented          in          Section~\ref{sec:liveness}.
Section~\ref{sec:computing} describes  the formulation of  liveness as
grammars. We also give a proof  of undecidability of such grammars and
show     how     they     can     be     approximated     by     DFAs.
Section~\ref{sec:GC-scheme}   discusses   details   of   the   garbage
collector, in particular, the use of  liveness DFAs for GC.  We report
our experimental  results in Section~\ref{sec:experiments}  along with
some  observations.  Section~\ref{sec:relatedwork}  discusses previous
work  related  to  GC and  liveness  and  Section~\ref{sec:conclusion}
discusses possible extensions and concludes the paper.
\section{The target language---syntax and semantics}
\label{sec:defs}
Figure~\ref{fig:lang-syntax} describes the syntax  of our language. It
is a first-order language with lazy semantics. Programs are restricted
to        be        in        Administrative        Normal        Form
(ANF)~\cite{chakravarty03perspective}, where  all actual  parameters to
functions  are  variables.  While  this  restriction  does not  affect
expressibility,  this form  has  the benefit  of  making explicit  the
creation of closures through the $\LET$ construct.   We
  assume that  $\LET$s  in  our  language  are  non-recursive;  in  the
  expression $\LET\,\, x  \leftarrow s\,\, \IN\,\, e$,  $x$ should not
  occur in  $s$. The restriction of  \LET\ to a single  definition is
for ease  of exposition---generalization to multiple  definitions does
not add conceptual difficulties.  We further restrict each variable in
a program  to be  distinct, so that  no scope  shadowing occurs; this
simplifies reasoning.

@@

A notion  that generalizes liveness  is {\em demand}.   While liveness
gives the patterns of future uses of a variable, demand represents the
future uses of the value of an expression.  The demand on an expression
$e$   is   also   a   set    of   access   paths---the   subset   of
$\{\acar,\acdr\}^\ast$ which the  context of $e$ may  explore of $e$'s
result.   To see  the need  for demands,  consider the expression
$(\LET~x\leftarrow  (\CDR~y)~\IN~   (\SRETURN~x))$.   Assume   that  the
context of this expression places  the demand $\lbrace \epsilon, \acar
\rbrace$. Since the  value of the expression is the  value of $x$, the
demand translates to the liveness  $[x \mapsto \lbrace \epsilon, \acar
  \rbrace]$.   Due  to  the  \LET\   definition  which  binds  $x$  to
$(\CDR~y)$, the liveness of $x$  now becomes the demand on $(\CDR~y)$.
This, in  turn, generates  the liveness $\lbrace  y.\epsilon, y.\acdr,
y.\acdr\acar  \rbrace$.   These are  the $y$-rooted  accesses
required to  explore $\lbrace  \epsilon, \acar  \rbrace$ paths  of the
result of $(\CDR~y)$.
%% As an  analogy with classical  (strong) liveness analysis,  $y$ and
%% $z$ are live at the entry $\pi: x:=y+z$, if and only if $x$ is live
%% at  exit  of $\pi$.   In  our  terminology, the  liveness  $\lbrace
%% \epsilon\rbrace $ of $x$ at the  exit from $\pi$ becomes the demand
%% on  $y+z$,  and  this,  in turn  generates  the  liveness  $\lbrace
%% y.\epsilon, z.\epsilon \rbrace$ at the entry of $\pi$.

>>>>

\begin{abstract} 
We consider  the problem of reducing  the memory required to  run lazy
first-order functional programs. Our approach is to analyze
programs  for liveness  of  heap-allocated data.   The  result of  the
analysis is  used to preserve  only live data---a subset  of reachable
data---during garbage  collection.  The result  is an increase  in the
garbage reclaimed and  a  reduction in  the peak memory requirement 
of programs.   Whereas this  technique has already  been shown  to yield
benefits for  eager first-order  languages, the  lack of  a statically
determinable execution  order and  the presence  of closures  pose new
challenges  for lazy  languages.  These  require changes  both in  the
liveness analysis itself and in the design of the garbage collector.

@@

Whereas  the runtime  system  of most  functional  languages includes  a
garbage collector to reclaim memory efficiently, empirical studies on
Scheme~\cite{karkare06effectiveness}  and  Haskell~\cite{rojemo96lag}
programs have shown that garbage  collectors leave uncollected a large
number of  memory objects that are  reachable but not live  (here {\em
  live} means the  object may potentially be used by  the program at a
later stage).  This results in unnecessary memory retention.

@@

Whereas  the idea  of  using static  analysis  to improve  memory
  utilization  has  been  shown  to   be  effective  for  {\em  eager}
  languages~\cite{asati14lgc,       HofmannJ03,       inoue88analysis,
    lee05static}, a straightforward extension  of the technique is not
  possible  for  lazy  languages,  where  heap-allocated  objects  may
  include   closures    (runtime   representations    of   unevaluated
  expressions).
%%\todelete{Designing  a  liveness  based  garbage
%%  collector   for  lazy   languages  poses   significant  challenges.}
Firstly, since  data is made  live by  evaluation of closures,  and in
lazy languages  the place in  the program where this  evaluation takes
place cannot  be statically determined, laziness  complicates liveness
analysis itself.  Moreover, for liveness-based  GC to be effective, we
need to  extend it  to closures  apart from  evaluated data.   Since a
closure can escape the scope in which  it was created, it has to carry
the  liveness  information  of   its  free  variables.   As  execution
progresses  and possible  future uses  are eliminated,  we update  the
liveness  information  in  a  closure with  a  more  precise  version.
For these  reasons,  the  garbage   collector  also  becomes
significantly more  complicated than a liveness-based  collector for an
eager language.

@@

Section~\ref{sec:defs}  introduces  the   syntax  of  the  programming
language considered  and gives a small-step  operational semantics for
it.  The liveness  analysis for this language and  its soundness proof
is          presented          in          Section~\ref{sec:liveness}.
Section~\ref{sec:computing} describes  the formulation of  liveness as
grammars. We also give a proof  of undecidability of such grammars and
show     how     they     can     be     approximated     by     DFAs.
Section~\ref{sec:GC-scheme}   discusses   details   of   the   garbage
collector, in particular, the use of  liveness DFAs for GC.  We report
our experimental  results in Section~\ref{sec:experiments}  along with
some  observations.  Section~\ref{sec:relatedwork}  discusses previous
work  related  to  GC and  liveness  and  Section~\ref{sec:conclusion}
discusses possible extensions and concludes the paper.
\section{The target language---syntax and semantics}
\label{sec:defs}
Figure~\ref{fig:lang-syntax} describes the syntax  of our language. It
is a first-order language with lazy semantics. Programs are restricted
to        be        in        Administrative        Normal        Form
(ANF)~\cite{chakravarty03perspective}, where  all actual  parameters to
functions  are  variables.  Whereas  this  restriction  does not  affect
expressibility,  this form  has  the benefit  of  making explicit  the
creation of closures through the $\LET$ construct.   We
  assume that  $\LET$s  in  our  language  are  non-recursive;  in  the
  expression $\LET\,\, x  \leftarrow s\,\, \IN\,\, e$,  $x$ should not
  occur in  $s$. The restriction of  \LET\ to a single  definition is
for ease  of exposition---generalization to multiple  definitions does
not add conceptual difficulties.  We further restrict each variable in
a program  to be  distinct, so that  no scope  shadowing occurs; this
simplifies reasoning.

@@

A notion  that generalizes liveness  is {\em demand}.   Whereas liveness
gives the patterns of future uses of a variable, demand represents the
future uses of the value of an expression.  The demand on an expression
$e$   is   also   a   set    of   access   paths---the   subset   of
$\{\acar,\acdr\}^\ast$ which the  context of $e$ may  explore of $e$'s
result.   To see  the need  for demands,  consider the expression
$(\LET~x\leftarrow  (\CDR~y)~\IN~   (\SRETURN~x))$.   Assume   that  the
context of this expression places  the demand $\lbrace \epsilon, \acar
\rbrace$. Since the  value of the expression is the  value of $x$, the
demand translates to the liveness  $[x \mapsto \lbrace \epsilon, \acar
  \rbrace]$.   Due  to  the  \LET\   definition  which  binds  $x$  to
$(\CDR~y)$, the liveness of $x$  now becomes the demand on $(\CDR~y)$.
This, in  turn, generates  the liveness $\lbrace  y.\epsilon, y.\acdr,
y.\acdr\acar  \rbrace$.   These are  the $y$-rooted  accesses
required to  explore $\lbrace  \epsilon, \acar  \rbrace$ paths  of the
result of $(\CDR~y)$.
%% As an  analogy with classical  (strong) liveness analysis,  $y$ and
%% $z$ are live at the entry $\pi: x:=y+z$, if and only if $x$ is live
%% at  exit  of $\pi$.   In  our  terminology, the  liveness  $\lbrace
%% \epsilon\rbrace $ of $x$ at the  exit from $\pi$ becomes the demand
%% on  $y+z$,  and  this,  in turn  generates  the  liveness  $\lbrace
%% y.\epsilon, z.\epsilon \rbrace$ at the entry of $\pi$.
